\begin{english}
    
\newpage
\thispagestyle{empty}

\begin{center}
    {\Large\textbf{Acknowledgments}}
\end{center}

\vspace{1cm}

All praise and thanks are due to Allah, who granted us the strength and perseverance to complete this work.

\vspace{0.7cm}

We are deeply indebted to our supervisor, \textbf{Dr. Qotadeh Saber Al-Jawazneh}, whose visionary guidance, scholarly insights, and unwavering support were the beacon that led us through this journey. His mentorship went beyond academic supervision; it shaped our critical thinking and refined our technical approach.

\vspace{0.7cm}

A special note of profound gratitude goes to \textbf{Dr. Anwar Katrawi}. His pivotal role in bridging the gap between academia and the clinical field was instrumental to our success. We thank him for his dedicated efforts in facilitating our access to medical resources and for closely monitoring our progress with keen interest.

\vspace{0.7cm}

We extend our most sincere appreciation to the administration and medical staff of \textbf{Zarqa Governmental Hospital}. We are particularly grateful for their openness to innovation and for providing the essential real-world environment to pilot ``MedEcho.'' Their cooperation in providing access to clinical data and allowing us to test our system within their facility transformed this project from a theoretical concept into a tangible, impactful reality.

\vspace{0.7cm}

To our families, who have been our silent pillars of strength, thank you for your endless patience, sacrifices, and belief in our potential.

\end{english}