% =========================================
% CHAPTER 7: Results and Discussion
% =========================================



\section{Overview}
\label{sec:results:overview}
This chapter presents the findings obtained from the development, testing, and pilot 
deployment of the MedEcho system. The results are analyzed based on technical performance 
(accuracy, latency, structured extraction) and clinical usability (workflow integration, 
user feedback). The discussion also evaluates how well the system addresses the original 
problem statement of reducing documentation burden while maintaining clinical quality.

% -------------------------------------
% 7.2 Technical Performance
% -------------------------------------
\section{Technical Performance Results}

\subsection{Transcription Accuracy (Speech-to-Text)}
\label{sec:results:accuracy}

The system’s ability to operate within a \textbf{code-switching environment} 
(Arabic + English) was a major success factor.

\subsubsection*{Findings}
OpenAI Whisper demonstrated strong robustness when transcribing Jordanian dialect 
mixed with English medical terminology.

\subsubsection*{Error Analysis}
\begin{itemize}
    \item Minor inaccuracies occurred with certain local drug brand names.
    \item The overall medical meaning remained intact.
    \item The Local Knowledge Layer (LKL) successfully corrected many context-specific terms.
\end{itemize}

\subsection{Processing Latency}
\label{sec:results:latency}

Latency measures the time between ending the audio recording and displaying the structured 
report.

\subsubsection*{Average Latency}
For a typical two-minute medical consultation:

\[
\textbf{15–20 seconds}
\]

\subsubsection*{Analysis}
Pilot testing confirmed this delay is acceptable since it aligns with the natural transition 
between physical examination and reporting review.

\subsection{Structured Data Extraction (OSCE Compliance)}
\label{sec:results:osce}

The system was tested for its ability to extract OSCE-compliant fields.

\subsubsection*{Results}
\begin{itemize}
    \item The system correctly categorized symptoms and patient details into appropriate OSCE sections in \textbf{90\%} of test cases.
    \item Implemented \textbf{Negative Findings Logic}: when symptoms were denied, they were recorded explicitly rather than omitted.
\end{itemize}

This logic was specifically validated based on requirements from  
\textbf{Dr. Samer Abu Ghazaleh}.

% -------------------------------------
% 7.3 Clinical Validation Results
% -------------------------------------
\section{Clinical Validation Results (Field Testing)}
Field evaluation took place at Zarqa Government Hospital with real clinical staff.

\subsection{Workflow Integration}
\label{sec:results:workflow}

\subsubsection*{Outcome}
The introduction of the \textbf{Two-Phase Workflow} (Phase 1 Intake + Phase 2 Final Assessment) 
was highly effective.

\subsubsection*{Impact}
\begin{itemize}
    \item Mirrors clinicians’ natural cognitive progression.
    \item Prevents premature or incorrect AI-generated diagnoses.
    \item Ensures that final medical judgment remains with the doctor.
\end{itemize}

\subsection{User Experience (UX) Feedback}
\label{sec:results:ux}

Feedback was collected from senior consultants and medical residents.

\subsubsection*{Positive Feedback}
\begin{itemize}
    \item Simple and intuitive UI.
    \item “\textbf{One-Click Record}” was highlighted as crucial in overloaded clinics.
\end{itemize}

\subsubsection*{Critical Feedback and Improvements}
Doctors frequently omitted allergy information.  
To address this:

\begin{itemize}
    \item The system now prompts for allergies.
    \item A safety default is included when no allergy information is provided.
\end{itemize}

This ensures improved patient safety and documentation completeness.

% -------------------------------------
% 7.4 Discussion
% -------------------------------------
\section{Discussion}

\subsection{Interpretation of Results}
The results indicate that Large Language Models, when combined with a 
\textbf{Local Knowledge Layer (LKL)}, can be effectively grounded for clinical tasks.  
MedEcho successfully converts unstructured medical speech into a legal-grade structured 
report.

The accurate detection of \textbf{Negative Findings} represents a major advancement, 
distinguishing MedEcho from simple transcription tools.

\subsection{Comparison with Manual Documentation}
\label{sec:results:comparison}

\subsubsection*{Time Efficiency}
Typing a complete medical report manually typically takes:

\[
5\text{–}10 \text{ minutes}
\]

Using MedEcho:

\[
1\text{–}2 \text{ minutes (dictation + review)}
\]

This represents a potential \textbf{80\% reduction in documentation time}.

\subsubsection*{Completeness of Records}
The system ensures consistent inclusion of:

\begin{itemize}
    \item Social History
    \item Family History
    \item Allergy Status
    \item Key OSCE elements
\end{itemize}

This reduces the risk of missing essential medical details due to time pressure.

\subsection{Limitations and Future Work}
\label{sec:results:limitations}

While results are promising, several limitations remain:

\begin{itemize}
    \item \textbf{Internet dependency:} LLM processing currently requires API connectivity.
    \item \textbf{Offline Models:} Future work should explore local inference using 
          quantized models for on-premise hospital deployment.
    \item \textbf{Hospital Integration:} Planned support for integration with Jordan’s 
          national system (\textit{Hakeem}) to streamline PDF uploads.
\end{itemize}

% -------------------------------------
% 7.5 Conclusion
% -------------------------------------
\section{Conclusion}
\label{sec:results:conclusion}

The MedEcho system successfully demonstrated that an AI-powered, voice-driven 
documentation assistant is both technically feasible and clinically beneficial.  
Pilot testing results from Zarqa Government Hospital strongly indicate that MedEcho can:

\begin{itemize}
    \item Reduce physician burnout,
    \item Improve documentation quality,
    \item Enhance patient safety through structured and complete reporting.
\end{itemize}

These results validate MedEcho as a promising and practical innovation for real-world 
clinical environments.
